---
title: "Geospatial with R"
subtitle: "Handling vector data with `sf`"
execute:
  freeze: auto
---

This is my practice sections following [R as GIS for Economists](https://tmieno2.github.io/R-as-GIS-for-Economists-Quarto/).

### Basics

Load the packages:
```{r}
#| warning: false
 
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  sf, # vector data operations
  tidyverse, # data wrangling
  data.table, # data wrangling
  tmap, # make maps
  mapview, # create an interactive map
  patchwork, # arranging maps
  rmapshaper
)

```

```{r}
#| warning: false
 
#--- a dataset that comes with the sf package ---#
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"))

ggplot() +
  geom_sf(data = nc) +
  theme_void()
```

```{r}

head(nc)

```

`sfg`: `LINESTRING`

```{r}
#--- collection of points in a matrix form ---#
s1 <- rbind(c(2, 3), c(3, 4), c(3, 5), c(1, 5))

#--- see what s1 looks like ---#
s1

#--- create a "LINESTRING" ---#
a_linestring <- sf::st_linestring(s1)

#--- check the class ---#
class(a_linestring)

```

`sfg`: `POLYGON`

```{r}
#--- a hole within p1 ---#
p1 <- rbind(c(0, 0), c(3, 0), c(3, 2), c(2, 5), c(1, 3), c(0, 0))
p2 <- rbind(c(1, 1), c(1, 2), c(2, 2), c(1, 1))

#--- create a polygon with hole ---#
a_plygon_with_a_hole <- sf::st_polygon(list(p1, p2))

plot(a_plygon_with_a_hole)

```

Creating `sfc` object or `sf` object:

```{r}

#--- create an sfc ---#
sfc_ex <- sf::st_sfc(list(a_linestring, a_plygon_with_a_hole))

#--- create a data.frame ---#
df_ex <- data.frame(name = c("A", "B"))

#--- add the sfc as a column ---#
df_ex$geometry <- sfc_ex

#--- take a look ---#
df_ex
```

```{r}

#--- let R recognize the data frame as sf ---#
sf_ex <- sf::st_as_sf(df_ex)

#--- see what it looks like ---#
sf_ex
```

Read shapefile:

```{r}
#--- read a NE county boundary shapefile ---#
nc_loaded <- sf::st_read("Data/nc.shp")

```

Projection with different CRS:

```{r}
sf::st_crs(nc)

```

Change the CRS projection:

```{r}
#--- transform ---#
nc_wgs84 <- sf::st_transform(nc, 4326)

#--- check if the transformation was successful ---#
sf::st_crs(nc_wgs84)

#--- transform ---#
nc_utm17N <- sf::st_transform(nc_wgs84, 26917)

#--- check if the transformation was successful ---#
sf::st_crs(nc_utm17N)

```

```{r}
#--- transform ---#
nc_utm17N_2 <- sf::st_transform(nc_wgs84, sf::st_crs(nc_utm17N))

#--- check if the transformation was successful ---#
sf::st_crs(nc_utm17N_2)

```

Turning dataframe into sf object

```{r}
#--- read irrigation well registration data ---#
(
  wells <- readRDS("Data/well_registration.rds")
)

```

```{r}
#--- recognize it as an sf ---#
wells_sf <- sf::st_as_sf(wells, coords = c("longdd", "latdd")) %>% 
  st_set_crs(4269)

#--- take a look at the data ---#
head(wells_sf[, 1:5])

```

Conversion to and from `sp` objects

```{r}
#--- conversion ---#
wells_sp <- as(wells_sf, "Spatial")

#--- check the class ---#
class(wells_sp)

#--- revert back to sf ---#
wells_sf <- sf::st_as_sf(wells_sp)

#--- check the class ---#
class(wells_sf)
```

Reverting `sf` object back into dataframe

```{r}
#--- remove geometry ---#
wells_no_longer_sf <- sf::st_drop_geometry(wells_sf)

#--- take a look ---#
head(wells_no_longer_sf)

```

```{r}
#--- read wells location data ---#
urnrd_wells_sf <-
  readRDS("Data/urnrd_wells.rds") %>%
  #--- project to UTM 14N WGS 84 ---#
  sf::st_transform(32614)

#--- create a one-mile buffer around the wells ---#
wells_buffer <- sf::st_buffer(urnrd_wells_sf, dist = 1600)

head(wells_buffer)

ggplot() +
  geom_sf(data = urnrd_wells_sf, size = 0.6, color = "red") +
  geom_sf(data = wells_buffer, fill = NA) +
  theme_void()

```

```{r}
NE_counties <-
  readRDS("Data/NE_county_borders.rds") %>%
  dplyr::filter(NAME %in% c("Perkins", "Dundy", "Chase")) %>%
  sf::st_transform(32614)

#--- generate area by polygon ---#
(
  NE_counties <- dplyr::mutate(NE_counties, area = st_area(NE_counties))
)

```

```{r}
#--- create centroids ---#
(
  NE_centroids <- sf::st_centroid(NE_counties)
)

ggplot() +
  geom_sf(data = NE_counties) +
  geom_sf_text(data = NE_centroids, aes(label = NAME)) +
  theme_void()

```

```{r}
IL_counties <- sf::st_read("Data/IL_county_detailed.geojson")

IL_counties_mssimplified <- rmapshaper::ms_simplify(IL_counties, keep = 0.01)

Cook <- filter(IL_counties, NAME == "Cook County")

Cook_simplify <- sf::st_simplify(Cook, dTolerance = 1000)

ggplot() +
  geom_sf(data = Cook_simplify) +
  theme_void()

```

### Spatial Interactions of Vector Data: Subsetting and Joining

```{r}
#--- create points ---#
point_1 <- sf::st_point(c(2, 2))
point_2 <- sf::st_point(c(1, 1))
point_3 <- sf::st_point(c(1, 3))

#--- combine the points to make a single  sf of points ---#
(
points <- 
  list(point_1, point_2, point_3) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(point_name = c("point 1", "point 2", "point 3"))
)

```

```{r}
#--- create points ---#
line_1 <- sf::st_linestring(rbind(c(0, 0), c(2.5, 0.5)))
line_2 <- sf::st_linestring(rbind(c(1.5, 0.5), c(2.5, 2)))

#--- combine the points to make a single  sf of points ---#
(
lines <- 
  list(line_1, line_2) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(line_name = c("line 1", "line 2"))
)

```

```{r}
#--- create polygons ---#
polygon_1 <- sf::st_polygon(list(
  rbind(c(0, 0), c(2, 0), c(2, 2), c(0, 2), c(0, 0)) 
))

polygon_2 <- sf::st_polygon(list(
  rbind(c(0.5, 1.5), c(0.5, 3.5), c(2.5, 3.5), c(2.5, 1.5), c(0.5, 1.5)) 
))

polygon_3 <- sf::st_polygon(list(
  rbind(c(0.5, 2.5), c(0.5, 3.2), c(2.3, 3.2), c(2, 2), c(0.5, 2.5)) 
))

#--- combine the polygons to make an sf of polygons ---#
(
polygons <- 
  list(polygon_1, polygon_2, polygon_3) %>% 
  sf::st_sfc() %>% 
  sf::st_as_sf() %>% 
  dplyr::mutate(polygon_name = c("polygon 1", "polygon 2", "polygon 3"))
)

```

```{r}
ggplot() +
  geom_sf(data = polygons, aes(fill = polygon_name), alpha = 0.3) +
  scale_fill_discrete(name = "Polygons") +
  geom_sf(data = lines, aes(color = line_name)) +
  scale_color_discrete(name = "Lines") + 
  geom_sf(data = points, aes(shape = point_name), size = 4) +
  scale_shape_discrete(name = "Points")  

```

```{r}
sf::st_intersects(points, polygons)

```

```{r}
set.seed(38424738)

points_set_1 <-
  lapply(1:5, function(x) sf::st_point(runif(2))) %>% 
  sf::st_sfc() %>% sf::st_as_sf() %>% 
  dplyr::mutate(id = 1:nrow(.))

points_set_2 <-
  lapply(1:5, function(x) sf::st_point(runif(2))) %>% 
  sf::st_sfc() %>% sf::st_as_sf() %>% 
  dplyr::mutate(id = 1:nrow(.))

```

```{r}
sf::st_is_within_distance(points_set_1, points_set_2, dist = 0.2)

```

```{r}
#--- Kansas county borders ---#
KS_counties <-
  readRDS("Data/KS_county_borders.rds") %>%
  sf::st_transform(32614)

#--- High-Plains Aquifer boundary ---#
hpa <- 
  sf::st_read("Data/hp_bound2010.shp") %>%
  .[1, ] %>%
  sf::st_transform(st_crs(KS_counties))

#--- all the irrigation wells in KS ---#
KS_wells <- 
  readRDS("Data/Kansas_wells.rds") %>%
  sf::st_transform(st_crs(KS_counties))

#--- US railroads in the Mid West region ---#
rail_roads_mw <- sf::st_read("Data/mw_railroads.geojson")

```

```{r}
hpa_cropped_to_KS <- sf::st_crop(hpa, KS_wells)

```

```{r}
counties_intersecting_hpa <- KS_counties[hpa_cropped_to_KS, ]

ggplot() +
  geom_sf(data = counties_intersecting_hpa) +
  geom_sf(data = hpa_cropped_to_KS, fill = "blue", alpha = 0.3) +
  theme_void()

```

```{r}
counties_intersecting_hpa

# sf_1[sf_2, op = topological_relation_type]

```

```{r}
KS_corn_price <-
  KS_counties %>%
  dplyr::mutate(corn_price = seq(3.2, 3.9, length = nrow(.))) %>%
  dplyr::select(COUNTYFP, corn_price)

KS_corn_price

```

```{r}
(
  KS_wells_County <- sf::st_join(KS_wells, KS_corn_price)
)

```

```{r}
KS_wells %>% aggregate(KS_counties, FUN = sum)

```

```{r}
# sf::st_join(sf_1, sf_2, join = \(x, y) st_is_within_distance(x, y, dist = 5))

# st_intersection() => does cropped join

```